---
title: "U.S. Department of Education Proficiency Data Dashboard"
author: "Juyuan Li"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggthemes)
library(patchwork)
library(ggrepel)
```


## Choropleth map for Math Assessment data from 2020 to 2021

```{r}
#Input Math Assessment data from 2020 to 2021 and filter
math_achievement21 <- read_csv("https://www2.ed.gov/about/inits/ed/edfacts/data-files/math-achievement-lea-sy2020-21.csv")
```

```{r, warning=FALSE}
#Filter the data
math_achievement21 <- math_achievement21 %>%
  filter(CATEGORY == "ALL", 
         GRADE == "00", 
         NUMVALID > 1000,
         PCTPROF != "LE1")

#calculate the mean value of math assessment in each state
state_means_math_assess21 <- math_achievement21 %>%
  group_by(STNAM) %>%
  summarize(mean_mathassess = mean(as.numeric(PCTPROF))) %>%
  arrange(desc(mean_mathassess))

math_table <- rbind(head(state_means_math_assess21, n = 3), tail(state_means_math_assess21, n = 3))

# Merge the dataset with the us map data
us_map <- map_data("state")

glimpse(us_map)

us_map_math_assess21 <- state_means_math_assess21 %>%
  mutate(litte_state = str_to_lower(STNAM)) %>%
  left_join(us_map, by=c("litte_state" = "region"))

## So I have my math map
p_math_map_21 <- ggplot(us_map_math_assess21, aes(x=long, y=lat)) +
  geom_polygon(aes(fill=`mean_mathassess`, group=group), color="gray30") + 
  coord_map() + 
  theme_map() +
  theme(legend.position = "bottom") + 
  labs(fill="Percentage\nof Proficent Student",
       title = "Math Assessment Results by State",
       subtitle="From 2020 to 2021") +
  scale_fill_gradient2(low = "gray99", high = "steelblue3")
p_math_map_21

```

## Choropleth map for Reading/Language Arts Assessment data from 2020 to 2021
```{r, warning=FALSE}
#Input Reading and language Assessment data from 2020 to 2021
rla_achievement21 <- read_csv("https://www2.ed.gov/about/inits/ed/edfacts/data-files/rla-achievement-lea-sy2020-21.csv")
```

```{r, warning=FALSE}
#Filter the data
rla_achievement21 <- rla_achievement21 %>%
  filter(CATEGORY == "ALL", 
         GRADE == "00", 
         NUMVALID > 1000)

#Similarly, I calculate the mean value of rla assessment results in each state
state_means_rla_assess21 <- rla_achievement21 %>%
  group_by(STNAM) %>%
  summarize(mean_rlaassess = mean(as.numeric(PCTPROF))) %>%
  arrange(desc(mean_rlaassess))

rla_table <- rbind(head(state_means_rla_assess21, n = 3), 
                   tail(state_means_rla_assess21, n = 3))

#Similarly, I merge the dataset with the us map data
us_map_rla_assess21 <- state_means_rla_assess21 %>%
  mutate(litte_state = str_to_lower(STNAM)) %>%
  left_join(us_map, by=c("litte_state" = "region"))

## So I have my rla map
p_rla_map_21 <- ggplot(us_map_rla_assess21, aes(x=long, y=lat)) +
  geom_polygon(aes(fill=`mean_rlaassess`, group=group), color="gray30") + 
  #  geom_text(data=county_centers, aes(label=County) ) + 
  coord_map() + 
  theme_map() +
  theme(legend.position = "bottom") + 
  labs(fill="Percentage\nof Proficent Student",
       title = "Reading/Language Arts Assessment Results by State",
       subtitle="From 2020 to 2021") +
  scale_fill_gradient2(low = "gray99", high = "firebrick3")
p_rla_map_21

```

## Draft dashboard
Now I want to compare the two choropleth maps and see if I can find any relations
```{r, warning=FALSE}
my_dashboard <- (p_math_map_21 | p_rla_map_21)+
  plot_annotation(
    title = 'Achievement Results for State Assessments',
    caption = 'Source: U.S. Department of Education') +
  plot_layout(heights=c(2.5, 1.5))
my_dashboard
#It seems like the relationship between math and reading/language arts is positive.
#Now i want to do a relationship plot to see if my assumption is correct.
```


## Daft Relationship Plot
```{r, warning=FALSE}
#First, I identify those states who have best and worst math and rla assessment results
which.min(state_means_math_assess21$mean_mathassess)
which.max(state_means_math_assess21$mean_mathassess)
state_means_math_assess21$STNAM[c(1,49)]#"MARYLAND" "MAINE"

which.min(state_means_rla_assess21$mean_rlaassess)
which.max(state_means_rla_assess21$mean_rlaassess)
state_means_rla_assess21$STNAM[c(1,49)]#"OKLAHOMA" "MAINE" 
```

## Build the virtualization

```{r, warning=FALSE}
states_of_interest <- c(state_means_math_assess21$STNAM[c(1,49)],
                        state_means_rla_assess21$STNAM[c(1,49)],
                        "NEW MEXICO",
                        "ALABAMA",
                        "MARYLAND") ## These three states appear opposite results from the map 
states_of_interest <- unique(states_of_interest)

#Then, I generate a relationship plot with states of interests labeled.
#Merge math assessment results with reading/language arts assessment results
state_means_math_rla_assess21 <- left_join(state_means_math_assess21,state_means_rla_assess21,
                                           by=c("STNAM" = "STNAM"))

#Draw the relationship
p_math_rla_relate <- ggplot(state_means_math_rla_assess21, 
                             aes(x=mean_rlaassess, 
                                 y=mean_mathassess)) + 
  geom_smooth(method="lm", color="gray85", se=FALSE) + 
  geom_point(color="gray50") + 
  geom_text_repel(data=filter(state_means_math_rla_assess21, STNAM %in% states_of_interest), aes(label=STNAM) ) + 
  labs(x="Average Reading/Language Arts Assessment Results", y="Average Math Assessment Results",
       title="Relationship of Math & Reading/Language Arts") + 
  theme_minimal()
p_math_rla_relate

```


## Add participation into relationship plot
```{r, warning=FALSE}
#load participation data and filter
math_participation21 <- read_csv("https://www2.ed.gov/about/inits/ed/edfacts/data-files/math-participation-lea-sy2020-21.csv")
rla_participation21 <- read_csv("https://www2.ed.gov/about/inits/ed/edfacts/data-files/rla-participation-lea-sy2020-21.csv")
```

```{r, warning=FALSE}
math_participation21 <-math_participation21 %>%
  filter(CATEGORY == "ALL", 
       GRADE == "00", 
       NUMPART > 1000)

rla_participation21 <-rla_participation21 %>%
  filter(CATEGORY == "ALL", 
         GRADE == "00", 
         NUMPART > 1000)

#For math participation(similar to assessment results)
state_means_math_part21 <- math_participation21 %>%
  group_by(STNAM) %>%
  summarise(mean_mathpart = mean(as.numeric(PCTPART), na.rm = TRUE))%>%
  arrange(desc(mean_mathpart))

#For rla participation(similar to assessment results)
state_means_rla_part21 <- rla_participation21 %>%
  group_by(STNAM) %>%
  summarise(mean_rlapart = mean(as.numeric(PCTPART), na.rm = TRUE))%>%
  arrange(desc(mean_rlapart))

#merge the two participation data sets and calculate the mean
state_means_math_rla_part21 <- left_join(state_means_math_part21,state_means_rla_part21,
                                           by=c("STNAM" = "STNAM")) %>%
  mutate(average_part = (mean_mathpart + mean_rlapart) / 2)
#merge the participation dataset with assessment results
state_means_math_rla_assess_part21 <- left_join(state_means_math_rla_assess21,state_means_math_rla_part21,
                                         by=c("STNAM" = "STNAM"))

```

## Build the virtualization

```{r, warning=FALSE}
#Draw the relationship again! This time I introduce participation rate.
p_math_rla_relate_withpart <- ggplot(state_means_math_rla_assess_part21, 
                            aes(x=mean_rlaassess, 
                                y=mean_mathassess)) + 
  geom_point(color="moccasin", aes(size=average_part)) + 
  geom_smooth(method="lm", color="green3", se=FALSE) + 
  geom_point(color="brown4", data=filter(state_means_math_rla_assess21, STNAM %in% states_of_interest), aes(label=STNAM)) +
  scale_size_continuous(name = "Participation rate") + 
  geom_text_repel(data=filter(state_means_math_rla_assess21, STNAM %in% states_of_interest), aes(label=STNAM) ) + 
  labs(x="Average Reading/Language Arts Assessment Results", y="Average Math Assessment Results",
       title="Relationship of Math & Reading/Language Arts") + 
  theme_minimal()
p_math_rla_relate_withpart
```


## Redo choropleth maps and label states of interest.
```{r, warning=FALSE}
# estimate centers of states
state_centers <- us_map_math_assess21 %>% 
  filter(STNAM %in% states_of_interest) %>% 
  group_by(STNAM) %>%
  summarize_at(vars(long, lat), ~ mean(range(.)))

#Final math assessment map
p_math_map_21_withlabels <- ggplot(us_map_math_assess21, aes(x=long, y=lat)) +
  geom_polygon(aes(fill=`mean_mathassess`, group=group), color="gray30") +
  geom_text(data=state_centers, aes(x = long, y = lat, label=STNAM), fontface = "bold", size = 3) + 
  coord_map() + 
  theme_map() +
  theme(legend.position = "bottom") + 
  labs(fill="Percentage\nof Proficent Student",
       title = "Math Assessment Results by State",
       subtitle="From 2020 to 2021") +
  scale_fill_gradient2(low = "gray99", high = "steelblue3")
p_math_map_21_withlabels

#Final reading/language arts assessment map
p_rla_map_21_withlabels <- ggplot(us_map_rla_assess21, aes(x=long, y=lat)) +
  geom_polygon(aes(fill=`mean_rlaassess`, group=group), color="gray30") +
  geom_text(data=state_centers, aes(x = long, y = lat, label=STNAM), fontface = "bold", size = 3) + 
  coord_map() + 
  theme_map() +
  theme(legend.position = "bottom") + 
  labs(fill="Percentage\nof Proficent Student",
       title = "Reading/Language Arts Assessment Results by State",
       subtitle="From 2020 to 2021") +
  scale_fill_gradient2(low = "gray99", high = "firebrick3")
p_rla_map_21_withlabels

```


## Final dashboard
```{r, message=FALSE, warning=FALSE}
my_dashboard_final <- (p_math_map_21_withlabels | p_rla_map_21_withlabels)/p_math_rla_relate_withpart   +
  plot_annotation(title = 'State Assessments in Mathematics and Reading/Language Arts',
                  caption = 'Source: U.S. Department of Education') +
  plot_layout(heights=c(2, 1.5))
my_dashboard_final

ggsave(plot=my_dashboard_final, filename="StateAssessmentDashboard.png",
       width=12, height=7.5, dpi=300)
```

**The code is designed to determine the overall state-wide average math proficiency percentages and state-wide average rla proficiency percentages for school districts with at least 1000 valid entries and shows in the map plot as the data virtualization. The blue map represents the math score across the nation, and the red map represents the rla score. In these two maps, the special thing we find is New Mexico State and Alabama state has negative correlation base on the color, both of them have high rla scores but low math scores. All other states seem to have a positive correlation with math and rla score. Based on this information, I create a linear plot to check the relationship between the math score and rla score. Overall, math score has a positive correlation with rla score, meaning that at the higher math score the state got, the higher rla score the state would also have. This does make sense because mostly a student have higher test score require both perfect in math and language, or both science and art. There are also few students have problems that only good at math or language but bad at the other one, however these students are few numbers, and mostly students still got a positive correlation on math and language scores, and that reflects on the data. As an English-speaking country, if the education participation rate is not high, we can still expect that the language score does not become very low because students keep living in an English-speaking environment, but the math score could be dropped due to the low participation. Therefore I tried looked at the participation data, turns out for several states with weak or negative correlation that have high rla score with low math score(Alabama, Maryland, and New Mexico), only New Mexico have a really low participation rate. Overall, It does make sense that Maine state have both high math and rla score with high participation rate, and New Mexico have high rla score but struggle on math due to the low participation rate. Although Alabama, Maryland, and Oklahoma state have both low scores on math and rla with good participation rate that seems to violate the theory that participation rate could affect the scores(especially math scores), but when we look at the dots below the green/mean line, there are several smaller dots represents low participation rate. Therefore we can determine that when a state has low correlation between math and rla score(rla perform much better than math), we can expect to see a comparable lower participation rate than normal in this state, this also supports that the theory of low participation rate could affect math score more than rla score. Lastly, for Alabama, Maryland, and Oklahoma should consider and find more factors to determine the low scores with high participation rates, we can do the hypothesis like lack of good education resources or something, but the hypothesis needs sources, research to test and prove, and this would be what future study should work on.**
